Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-08 analysis/paper/article.qmd

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year. @tbl-show-basic-counts shows the breakdown by different job types, ranks, and tenure status. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

Our study of a decade of tenure-track job ads in archaeology in the US reveals diverse dynamics in the demand for specialisations in topics, methods and geographic focus, and in the instructions to applicants. While these dynamics are familiar to scholars actively seeking faculty positions, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe affected by a variety of factors. Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. This is at odds with previous work that has shown that contingent positions make up more than half of faculty positions [@colby2022annual], and that these numbers are increasing— between 2004 and 2019, for example, there was a 36% increase in the replacement of tenure-track positions by contingent positions [@the20222022]. This discrepancy in our data set may be due to the more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but only through local email lists and are filled by people close to the hiring department, such as recently graduated students. One possible explanation for our results is that non-tenure-track jobs are less frequently added to the Academic Jobs Wiki. The downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

>>>>

We collected data from `r total_number_of_ads_in_our_sample` ads for tenure-track jobs in archaeology posted during 2013-2023. We focus our analysis here on the `r total_number_of_us_ads_in_our_sample` ads for positions at US universities. @fig-show-basic-plots shows the count of ads for each year, where year refers to the year the job ad was posted. @tbl-show-basic-counts shows the breakdown by different job types, ranks, and tenure status. Assistant Professor jobs are consistently the most common title and rank of positions advertised, while open rank or full professor are the least frequent. The ratio of tenure-track to non-tenure-track positions is generally well above one; in other words, this data set is dominated by tenure-track positions. Only academic year 2013-2014 had more non-tenure-track positions than tenure-track positions, which was followed by an upward trend peaking at 2018-2019 and then declining again into the present.

@@

Our study of a decade of tenure-track job ads in archaeology in the US reveals diverse dynamics in the demand for specialisations in topics, methods and geographic focus, and in the instructions to applicants. While these dynamics are familiar to scholars actively seeking faculty positions, we believe this is the first time they have been quantified at such a large scale within archaeology. Trends in job ads reflect broader shifts in intellectual and practical priorities concerning archaeology, undergraduate education, and the process of hiring professors. The demand for archaeology faculty, indicated by the total number of tenure-track jobs, maybe affected by a variety of factors.

Overall, we found more tenure-track jobs advertised each year than non-tenure-track, with the exception of 2013-2014. This is at odds with previous work that has shown that contingent positions make up more than half of faculty positions [@colby2022annual], and that these numbers are increasing— between 2004 and 2019, for example, there was a 36% increase in the replacement of tenure-track positions by contingent positions [@the20222022]. This discrepancy in our data set may be due to several factors. The more limited circulation of advertising for short-term positions relative to advertising for tenure-track jobs. Many of these short-term positions are not advertised nationally, but only through local email lists and are filled by people close to the hiring department, such as recently graduated students. Another factor is bias in our data, with most Wiki users likely seeking a tenure-track job, non-tenure-track jobs are may have been less frequently added to the Academic Jobs Wiki because they were peripheral to the goal of most users. This bias limits the reliability of our results on the ratio of tenure-track to non-tenure-track.

The downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]
